Quantifying Infants' Everyday Restrained Experiences in the Home Using Wearable Inertial Sensors

### 2025-11-16 manuscript_restrained_classification validation.qmd
updated abstract and fixed typo

author-note:
  blank-lines-above-author-note: 1
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    # Acknowledge and cite data/materials to be shared.
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    # Example: This study was registered at ClinicalTrials.gov (Identifier NTC998877).
    study-registration: null
    data-sharing: null
    # Example: This article is based on data published in Pulaski (2017).
    # Example: This article is based on the dissertation completed by Graham (2018).    
    # Example: Sally Jones earns royalties from the sale of Test X.
    conflict-of-interest: null
    # Example: This study was supported by Grant A123 from the National Science Foundation.
    financial-support: null
    # Example: The authors are grateful for the technical assistance of Dr. X during the initial design and setup of our lab equipment.
    gratitude: null
    authorship-agreements: null
abstract: "Physical restraint—including being held, carried, and restrained in devices—is a common feature of infants’ everyday lives. This study developed and validated a machine learning model to detect infants’ restraint using wearable sensors movement data derived from an existing data set (Franchak, 2023). The model achieved high accuracy (89%) and substantial kappa agreement (0.73) compared with human-coded ground truth. The model presented a slight bias towards overestimating unrestrained periods compared to restrained periods, but the bias was mitigated when using a longer data-aggregating window length. The model also showed convergent validity by corroborating prior studies that showed an age-related decrease in infant restraint time. In short, the current study demonstrated the feasibility of using wearable sensors to quantify infants’ real-world restraint experiences, offering a new tool for studying how daily restraint influences early development."
keywords: [infant, restraint, everyday experiences, wearable sensors, machine learning]
bibliography: [references.bib]
format:
  apaquarto-docx: 
    floatsintext: false
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf: 
    # can be jou (journal), man (manuscript), stu (student), or doc (document)
    # for now, tables and figures do not render properly in jou mode. 
    # can be 10pt, 11pt, or 12pt
    # Integrate tables and figures in text
    # a4 paper if true, letter size if false
    # suppresses loading of the lmodern font package
    # Suppresses loading of the fontenc package
    # Suppress the title above the introduction
    # In jou mode, use times or pslatex instead of txfonts
    # In jou mode, use Computer Modern font instead of times
    # In jou mode, cancels automatic stretching of tables to column width 
    # Uses Helvetica font in stu and man mode
    # In man and stu mode, neutralizes the \helvetica command
    # In man and stu mode, uses typewriter font
    # Puts draft watermark on first page
    # Puts draft watermark on all pages
    # Masks references that are marked as the author's own
    # Hides correspondence text
    documentmode: man
    fontsize: 12pt
    floatsintext: false
    a4paper: false
    nolmodern: false
    nofontenc: false
    donotrepeattitle: false
    notxfonts: false
    notimes: false
    notab: false
    helv: false
    nosf: false
    tt: false
    draftfirst: false
    draftall: false
    mask: false
    journal: null
    volume: null
    course: null
    professor: null
    duedate: null
    nocorrespondence: false
editor_options: 
  chunk_output_type: console
---

@@

Physical restraint is a common feature in infants’ everyday lives. Infants are routinely held or carried by caregivers, and infants are often restrained in devices such as highchairs, walkers, and strollers [@Birken2015s]. As an externally regulated aspect of infants’ daily life, restraint shapes infants’ experiences based on caregivers’ decisions. Restraint limits infants’ physical activities and therefore hinders infants’ motor development [e.g., @Pin2007z]. Restraint also frames infants’ accessible physical environment by altering experiences that are relevant for perceptual and cognitive development. For example, restraint is associated with fewer object holding experiences [@Franchak2024l], reduced amount and quality of language input [@Malachowski2023a], and increased opportunities for joint attention [@Karasik2025h]. Therefore, characterizing infants’ restrained experiences helps profile infants’ daily learning contexts and situate their motor, perceptual, and cognitive development within everyday experiences.

Previous studies have used a variety of methods to document infants’ restrained time in the home. Some used parent surveys, including questionnaires, diaries, or ecological momentary assessment [e.g., @Abbott2001u; @Karasik2018t; @Franchak2019o]. Some used video recording and human annotation [e.g., @Springfield2025p; @Wang2025v]. Although surveys provide an overall profile of infants’ daily life, and video recordings provide continuous time series data, no method so far can simultaneously provide full-day and continuous data to account for infants’ restraint experiences. In the current study, we propose that wearable sensors can be a solution for recording full-day, continuous movement behavior, and machine learning algorithms can be an accurate and efficient way to automatically classify wearable sensor data.

Moreover, previous studies did not define restraint consistently. Some definitions only focused on caregiver restraint such as holding and carrying [e.g., @Airaksinen2024z; @Franchak2021k; @Franchak2024e; @Yao2019q], some only focused on seating devices [e.g., @Callahan1997s], some looked at different surfaces (e.g., lap, cradle, and seat) infants are on [e.g., @Graciosa2024o], some focused broadly on containment equipment [e.g., @Bartlett2003w; @Carson2022n; @Hesketh2015i; @Karasik2022a], some combined holding and carrying with seating equipment [@Springfield2025p], and others covered a wider range of restraint like furniture placement, seating devices and holding [@Franchak2024l; @Malachowski2023a; @Siddicky2021c]. The inconsistency in restraint definitions makes it difficult to compare results across studies. In the current study, we take a broader scope and consider any practices that limit infants’ freedom of movement as restraint. We do not pre-define a list of restraint practices but focus on infants’ real-time opportunities for movement by classifying restraint whenever infants cannot change body position or locomote freely. Quantifying both caregiver restraint (i.e., holding and carrying) and device restraint, we can get a whole picture of the external constraints that infants experience in daily life.

@@

Despite the inconsistency in definitions, prior work converges to show that caregiver and device restraints are common in infant daily life, and the time infants spend restrained changes over development. In terms of caregiver restraint, 2- to 6-month-old infants spent around 40% of their awake time being held or carried [@Malachowski2023a; @Siddicky2020f]. Older 10- to 13-month-olds spent around 19% of their awake time being held or carried [@Franchak2024l]. In terms of device restraint, 4- to 6-month-olds spent around 20% of awake time in seating devices, and 35% in placement with physical support [@Callahan1997s; @Malachowski2023a]. Older 10- to 13-month-olds spent around 27% of awake time in restraining furniture or devices [@Franchak2024l]. Additionally, a study that compared across ages also confirmed that infants’ overall restrained time decreased from 2 to 6 months of age [@Carson2022n].

@@

Video recording and human annotation overcomes the above drawbacks of survey-based methods. However, one of the limitations of video observation is being obtrusive. The presence of a researcher in the room may affect the caregiver’s and infant’s behaviors. Another limitation is that video recording and video annotation are both labor-intensive and time-consuming, which limits its use in large-scale or long-duration studies. Indeed, prior work that videotaped and manually annotated infants’ restraint had short durations of recording (e.g., half an hour in @Springfield2025p, and one hour in @Wang2025v). Within the short duration of recording, researchers tend to record the period when infants are awake and active. In contrast, periods when the infants are restrained, like seated in front of TV or in highchairs, would likely be skipped. Thus, video recording is subject to sampling bias and might not be representative of infants’ whole-day experiences.

@@

Detecting restrained periods might be more challenging than classifying body postures. First, infants may show similar body movements whether restrained or not, making restrained and unrestrained classes more difficult to differentiate. For example, infants might kick legs when they are lying without restraint and also when they are strapped in a highchair. Possible distinctions might be manifested in reduced range of motion or repetitive motion due to constraints, which requires high-precision sensors and sensitive models to detect. Second, there are lots of possible restraint forms infants can present—caregiver holding can be holding the infant up straight or holding in arms breastfeeding; seating restraint can happen in countless infant seats, which vary in affordances of moving [@Alghamdi2024d]. Accordingly, the wearable sensor system should be robust enough to generalize across heterogeneous restraint scenarios. Thus, the dual requirement of sensitivity and robustness is the key challenge for developing a reliable model to detect infant restraint in the home.

Strategies to address this untested classification lie in the flexibility of modeling approaches. One strategy is adjusting the window length within which movement data are aggregated. With a fixed sampling rate, larger windows provide more information for the model to leverage to predict the current category. Prior studies have used window lengths ranging from 2.3 to 4 seconds for infant posture category [@Airaksinen2020j; @Franchak2024e], and from 1.15 to 10 seconds for infant holding and carrying detection [@Airaksinen2024z]. Considering that transitions between restraint states are typically slower than transitions between postures, we set window lengths longer than 4 seconds. Theoretically, shorter windows have higher temporal resolution and are sensitive to abrupt changes, but less data might mean the model lacks context to make an accurate classification. In contrast, longer windows can incorporate more data but may lack temporal sensitivity or may introduce irrelevant data from heterogeneous movement within an event. Determining the optimal window length is therefore a key step for the model to reliably detect infants’ restrained periods.

@@

The data set is hosted on Databrary (Franchak, 2023) and has been used in other publications. The data set consists of two cohorts of infants: `r apa_num(age[1,2])` younger infants aged from 4 to 7 months of age, and `r apa_num(age[2,2])` older infants aged from 11 to 14 months of age. Among the total of `r apa_num(age[1,2]+age[2,2])` infants, 26 of them participated in the study for only once, while 39 of them participated from 2 to 4 times, resulting in a total of 142 sessions. Thirty-four of the infants are female and thirty-one are male. Caregivers reported the infants’ race as White (*n* = 23), Hispanic or Latinx (*n* = 8), African American (*n* = 2), Asian (*n* = 2), more than one (*n* = 15), or not fitting into the above races or chose not to answer (*n* = 15).

@@

Using the 1.5-hour videos, our team annotated the periods when the infant was restrained or unrestrained using Datavyu software (https://datavyu.org/). Restrained and unrestrained was defined as a binary category, meaning that all times were coded as one of the two categories. A period was coded as missing if the infant went out of view and was excluded from further analysis. Restraint was defined as when infants could not initiate or control their own movements, which included being held or carried by caregivers, or being restrained in devices that constrain them from changing body position. Unrestrained periods, in contrast, were when infants could freely change their body position or locomote. For example, an infant was considered restrained when strapped into a high chair seat, but not when they were placed in a crib and free to change position. Each video file was coded by two coders: a primary coder annotated the whole 1.5-hour video, and an independent reliability coder annotated the first 30 minutes of the video. Inter-rater reliability was calculated as the proportion of video frames where the two coders coded the same label. The human annotation reached a high agreement with an overall accuracy of `r apa_num(stats_human[1,1]*100)`%, ranging from `r apa_num(stats_human[1,2]*100)`% to `r apa_num(stats_human[1,3]*100)`%, and an average kappa of `r apa_num(stats_human[1,4])`, ranging from `r apa_num(stats_human[1,5])` to `r apa_num(stats_human[1,6])`.

@@

We aggregated raw IMU data into motion features within sliding windows of *n* seconds, where the window length *n* varied from 4 s to 16 s and 30 s. For each window, we computed statistical features—mean, median, minimum, 25th percentile, 75th percentile, maximum, skewness, kurtosis, standard deviation, and sum—for all six signals across the four sensors, resulting in 240 features. We then added 196 additional features based on cross-sensor and cross-orientation metrics such as correlations and pairwise differences, resulting in 436 features per window.

We started the windowing process from the synchronization point and slid the window every *n*/2 seconds. Consequently, each time point spaced every *n*/2 seconds contains 436 motion features, aggregated from the *n*/2 seconds before and after that point. For model training and testing, we included only time windows where the infant remained in a single status, so that each window could be assigned one unambiguous label.

The third step was to train and validate supervised machine learning models that classify the motion features into restrained status. We applied the random forest algorithm (Breiman, 2001) in modeling using the *DecisionTree* package (Sadeghi et al., 2022) in Julia. We used a leave-one-out cross-validation approach, where we held out one session at each iteration as testing set and trained the model using the other sessions,to evaluate how well the model generalized to unseen data. After validation, an optimal window length was picked[,]{.underline} and a final model was trained using all the human annotations to process whole-day sensor data of all sessions.

## Data Sharing

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). Model evaluation, statistical analyses, and this reproducible manuscript were conducted in RStudio environment (R version 4.5.0; RStudio version 2025.5.0.496; RStudioTeam, 2024) using the R packages tidyverse [@tidyverse], flextable [@flextable], and papaja [@papaja], as well as the apa quarto template (Schneider, n.d.). All human annotation results and processing scripts are publicly shared on GitHub (https://github.com/Hanzhi-W/ms_restrained_classification-validation.git).

@@

We first compared the model’s performance to the human annotation of the 1.5-hour video recorded period across all 142 sessions using leave-one-out cross-validation approach. We reported performance metrics including accuracy, kappa, sensitivity, positive predictive value, F1, and the correlation of overall estimate time of restraint between the model and human annotation. We also examined prediction biases model and how they vary across window lengths. Second, we applied the model with the optimal window length to process infants’ full-day sensor data and examined the model’s convergent validity by replicating a previously-reported effect: that restraint time decreases with age [@Carson2022n].

@@

```{r tbl-indexbetweenmodels}
#| tbl-cap: Performance metrics according to window length
#| ft.align: left
#| apa-note: F, df, and p value indicate one-way ANOVA on each metric comparing across the three window lengths.

@@

```{r fig-prevalencegap}
#| fig-cap: Restraint prevalence 
#| fig-width: 8
#| fig-height: 4
#| apa-note: Prevalence is defined as the proportion of windows predicted (or annotated) as restrained among all windows. (A) shows the correlation of prevalence between model prediction and human annotation. The black dashed identity line indicates perfect agreement. Each circular symbol represents one session for a specific window length (denoted by color). (B) shows the difference of prevalence between human annotation and model prediction (i.e., prediction error). A negative prediction error indicates the model estimated less amount of time infants were restrained compared to human annotation. The horizontal line in the box denotes the median; the boarder lines of the box denotes the first (Q1) quartile and the third (Q3) quartile; and the lines extended denotes the min and max data points within the range of 1.5 IQR (i.e., Q3-Q1).

@@

@fig-prevalencegap A shows that, at session level, all window lengths provided reliable estimate of restrained time. We defined prevalence as the proportion of windows classified as restrained among all windows for each session. The prevalence predicted by the model was highly correlated with the prevalence annotated by human coders (*r* = `r apa_num(table_prev[1,4])` for 4 s, *r* = `r apa_num(table_prev[2,4])` for 16 s, and *r* = `r apa_num(table_prev[3,4])` for 30 s).

@@

Despite the overall good accuracy, models using all three window lengths were biased towards predicted that infants were unrestrained rather than restrained. @fig-cibtwctg shows the difference of a metric’s value (e.g., sensitivity, PPV) between restrained class and unrestrained class within a single window length. Larger difference scores (away from 0) indicate larger asymmetry in predicting one category rather than the other. @fig-cibtwctg A shows that the sensitivity difference for 4 s window was `r apa_num(table_btw_ctg[1,3])`, 95% bootstrapped CI \[`r apa_num(table_btw_ctg[1,4])`, `r apa_num(table_btw_ctg[1,5])`\], for 16 s window was `r apa_num(table_btw_ctg[3,3])`, 95% CI \[`r apa_num(table_btw_ctg[3,4])`, `r apa_num(table_btw_ctg[3,5])`\], and for 30 s window was `r apa_num(table_btw_ctg[5,3])`, 95% CI \[`r apa_num(table_btw_ctg[5,4])`, `r apa_num(table_btw_ctg[5,5])`\]. The negative differences of sensitivity across all window lengths indicate that the model was more sensitive in detecting the unrestrained class compared to restrained class regardless of window lengths. @fig-cibtwctg B shows that the PPV difference for 4 s window was `r apa_num(table_btw_ctg[2,3])`, 95% CI \[`r apa_num(table_btw_ctg[2,4])`, `r apa_num(table_btw_ctg[2,5])`\], for 16 s window was `r apa_num(table_btw_ctg[4,3])`, 95% CI \[`r apa_num(table_btw_ctg[4,4])`, `r apa_num(table_btw_ctg[4,5])`\], and for 30 s window was `r apa_num(table_btw_ctg[6,3])`, 95% CI \[`r apa_num(table_btw_ctg[6,4])`, `r apa_num(table_btw_ctg[6,5])`\]. The positive differences of PPV across all window lengths indicate that the model’s prediction of restrained periods was more credible compared to its prediction of unrestrained periods. Notably, while none of the window lengths were free from bias, the 30 s window presented smaller offsets of metric values between restrained and unrestrained detection.

The prediction bias was also evident when examining the model’s overall estimate of restrained time and its difference from human annotation. @fig-prevalencegap B shows that the model predictions consistently underestimated overall restrained time compared to human annotation, despite the high correlation between the two. We defined prediction error as the model predicted prevalence (in % units) subtracted by the human annotated prevalence of restrained periods (in % units). The 4 s window had a prediction error of `r apa_num(table_prev[1,5]*100)`%, 95% CI \[`r apa_num(table_prev[1,6]*100)`%, `r apa_num(table_prev[1,7]*100)`%\], the 16 s window of `r apa_num(table_prev[2,5]*100)`%, 95% CI \[`r apa_num(table_prev[2,6]*100)`%, `r apa_num(table_prev[2,7]*100)`%\], and the 30 s window of `r apa_num(table_prev[3,5]*100)`%, 95% CI \[`r apa_num(table_prev[3,6]*100)`%, `r apa_num(table_prev[3,7]*100)`%\].

Across the metrics of sensitivity, PPV, and prediction error, 30 s turned out to be the best window length out of the three, which brought about a balanced performance across classes.

@@

>>>>

author-note:
  blank-lines-above-author-note: 1
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    # Acknowledge and cite data/materials to be shared.
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    # Example: This study was registered at ClinicalTrials.gov (Identifier NTC998877).
    study-registration: null
    data-sharing: null
    # Example: This article is based on data published in Pulaski (2017).
    # Example: This article is based on the dissertation completed by Graham (2018).    
    # Example: Sally Jones earns royalties from the sale of Test X.
    conflict-of-interest: null
    # Example: This study was supported by Grant A123 from the National Science Foundation.
    financial-support: null
    # Example: The authors are grateful for the technical assistance of Dr. X during the initial design and setup of our lab equipment.
    gratitude: null
    authorship-agreements: null
abstract: "Physical restraint—including being held, carried, and restrained in devices—is a common feature of infants’ everyday lives. However, previous survey-based and video-based methods cannot simultaneously provide continuous, full-day accounts of infant restraint experiences. This study developed and validated a machine learning model to quantify infants’ restraint time moment-to-moment across the full day in the home environment using wearable sensors movement data derived from an existing data set. The data set includes 142 home-visit sessions from 65 infants, with 29 younger infants aged 4-7 months, and 36 older infants aged 11-14 months. We annotated infants’ restrained states in the first 1.5-hour video recordings of each session as ground truth labels. The supervised machine-learning model achieved high accuracy (89%) and substantial kappa agreement (κ = .73) compared with human-coded ground truth. The model presented a slight bias towards overestimating unrestrained periods compared to restrained periods, but the bias was mitigated when using a longer data-aggregating window length. The model also showed convergent validity by corroborating prior studies that showed an age-related decrease in infant overall restraint time throughout the day. In short, the current study demonstrated the feasibility of using wearable sensors to quantify infants’ real-world restraint experiences, offering a new tool for studying how daily restraint influences early development."
keywords: [infant, restraint, everyday experiences, wearable sensors, machine learning]
bibliography: [references.bib]
format:
  apaquarto-docx: 
    floatsintext: false
  apaquarto-html: default
  apaquarto-typst: default
  apaquarto-pdf: 
    # can be jou (journal), man (manuscript), stu (student), or doc (document)
    # for now, tables and figures do not render properly in jou mode. 
    # can be 10pt, 11pt, or 12pt
    # Integrate tables and figures in text
    # a4 paper if true, letter size if false
    # suppresses loading of the lmodern font package
    # Suppresses loading of the fontenc package
    # Suppress the title above the introduction
    # In jou mode, use times or pslatex instead of txfonts
    # In jou mode, use Computer Modern font instead of times
    # In jou mode, cancels automatic stretching of tables to column width 
    # Uses Helvetica font in stu and man mode
    # In man and stu mode, neutralizes the \helvetica command
    # In man and stu mode, uses typewriter font
    # Puts draft watermark on first page
    # Puts draft watermark on all pages
    # Masks references that are marked as the author's own
    # Hides correspondence text
    documentmode: man
    fontsize: 12pt
    floatsintext: false
    a4paper: false
    nolmodern: false
    nofontenc: false
    donotrepeattitle: false
    notxfonts: false
    notimes: false
    notab: false
    helv: false
    nosf: false
    tt: false
    draftfirst: false
    draftall: false
    mask: false
    journal: null
    volume: null
    course: null
    professor: null
    duedate: null
    nocorrespondence: false
editor_options: 
  chunk_output_type: console
---

@@

Physical restraint is a common feature in infants’ everyday lives. Infants are routinely held or carried by caregivers, and infants are often restrained in devices such as highchairs, walkers, and strollers [@Birken2015s]. As an externally regulated aspect of infants’ daily life, restraint shapes infants’ experiences based on caregivers’ decisions--caregivers, not infants, choose when infants are restrained. Restraint limits infants’ physical activities and might hinder infants’ motor development [e.g., @Pin2007z]. Restraint also alters experiences that are relevant for perceptual and cognitive development by determining what is accessible in the physical world. For example, restraint is associated with fewer object holding experiences [@Franchak2024l], reduced amount and quality of language input [@Malachowski2023a], and increased opportunities for joint attention [@Karasik2025h]. Therefore, characterizing infants’ restrained experiences helps profile infants’ daily learning contexts and situate their motor, perceptual, and cognitive development within everyday experiences.

Previous studies have used a variety of methods to document infants’ restrained time in the home. Some used parent surveys, including questionnaires, diaries, or ecological momentary assessment [e.g., @Abbott2001u; @Karasik2018t; @Franchak2019o]. Some used video recording and human annotation [e.g., @Springfield2025p; @Wang2025v]. Although surveys provide an overall profile of infants’ daily life and video recordings provide continuous time series data, neither method can simultaneously provide full-day and continuous data to account for infants’ restraint experiences. In the current study, we propose that wearable sensors can be a solution for recording full-day, continuous movement behavior, and machine learning algorithms can be an accurate and efficient way to automatically classify wearable sensor data.

@@

Despite different approaches, prior works converge to show that restraint is common in infant daily life, and that the time infants spend restrained changes over development. In terms of caregiver restraint, 2- to 6-month-old infants spent around 40% of their awake time being held or carried [@Malachowski2023a; @Siddicky2020f]. Older 10- to 13-month-olds spent around 19% of their awake time being held or carried [@Franchak2024l]. In terms of device restraint, 4- to 6-month-olds spent around 20% of awake time in seating devices, and 35% in placement with physical support [@Callahan1997s; @Malachowski2023a]. Older 10- to 13-month-olds spent around 27% of awake time in restraining furniture or devices [@Franchak2024l]. Additionally, a study that compared across ages also confirmed that infants’ overall restrained time decreased from 2 to 6 months of age [@Carson2022n].

Many other studies have documented some portion of infant restraint time, but they differed in restraint definitions, making it difficult to compare results across studies. Some definitions only focused on caregiver restraint such as holding and carrying [e.g., @Airaksinen2024z; @Franchak2021k; @Franchak2024e; @Yao2019q], some only focused on seating devices [e.g., @Callahan1997s], some looked at different surfaces (e.g., lap, cradle, and seat) infants are on [e.g., @Graciosa2024o], some focused broadly on containment equipment [e.g., @Bartlett2003w; @Carson2022n; @Hesketh2015i; @Karasik2022a], some combined holding and carrying with seating equipment [@Springfield2025p], and others covered a wider range of restraint like furniture placement, seating devices and holding [@Franchak2024l; @Malachowski2023a; @Siddicky2021c].

In the current study, we take a broader scope and consider any practices that limit infants’ freedom of movement as restraint. We do not pre-define a list of restraint practices but focus on infants’ real-time opportunities for movement by classifying restraint whenever infants cannot change body position or locomote freely. Quantifying both caregiver restraint (i.e., holding and carrying) and device restraint, we can get a whole picture of the external constraints that infants experience in daily life.

@@

Video recording and human annotation overcomes the above drawbacks of survey-based methods, but video is subject to other limitations. One disadvantage of video observation is being obtrusive. The presence of a researcher in the room may affect the caregiver’s and infant’s behaviors. Another limitation is that video recording and video annotation are both labor-intensive and time-consuming, which limits its use in large-scale or long-duration studies. Indeed, prior work that videotaped and manually annotated infants’ restraint had short durations of recording (e.g., half an hour in @Springfield2025p, and one hour in @Wang2025v). Within the short duration of recording, researchers tend to record the period when infants are awake and active. In contrast, periods when the infants are restrained, like seated in front of TV or in highchairs, would likely be skipped. Thus, video recording is subject to sampling bias and might not be representative of infants’ whole-day experiences.

@@

Detecting restrained periods might be more challenging than classifying body postures. First, infants may show similar body movements whether restrained or not, making restrained and unrestrained classes more difficult to differentiate. For example, infants might kick legs when they are lying without restraint and also when they are strapped in a highchair. Possible distinctions might be manifested in reduced range of motion or repetitive motion due to constraints, which requires high-precision sensors and sensitive models to detect. Second, there are lots of possible restraint forms infants can present--a caregiver can hold an infant up straight or in arms while breastfeeding; seating restraint can happen in countless infant seats, which vary in affordances of moving [@Alghamdi2024d]. Accordingly, the wearable sensor system should be robust enough to generalize across heterogeneous restraint scenarios. Thus, the dual requirement of sensitivity and robustness is the key challenge for developing a reliable model to detect infant restraint in the home.

Strategies to address this untested classification lie in the flexibility of modeling approaches. One strategy is adjusting the window length within which movement data are aggregated. With a fixed sampling rate, larger windows provide more information for the model to leverage to predict the current category. Prior studies have used window lengths ranging from 2.3 to 4 seconds for infant posture detection [@Airaksinen2020j; @Franchak2024e], and from 1.15 to 10 seconds for infant holding and carrying detection [@Airaksinen2024z]. Considering that transitions between restraint states are typically slower than transitions between postures, we set window lengths longer than 4 seconds. Theoretically, shorter windows have higher temporal resolution and are sensitive to abrupt changes, but fewer samples in a shorter window might mean the model lacks context to make an accurate classification. In contrast, longer windows can incorporate more data but may lack temporal sensitivity or may introduce irrelevant data from heterogeneous movement within an event. Determining the optimal window length is therefore a key step for the model to reliably detect infants’ restrained periods.

@@

The data set is hosted on Databrary (Franchak, 2023). The data set consists of two cohorts of infants: `r apa_num(age[1,2])` younger infants aged 4 to 7 months, and `r apa_num(age[2,2])` older infants aged 11 to 14 months. Among the total of `r apa_num(age[1,2]+age[2,2])` infants, 26 of them participated in the study for only once, while 39 of them participated from 2 to 4 times, resulting in a total of 142 sessions. Thirty-four of the infants are female and thirty-one are male. Caregivers reported the infants’ race as White (*n* = 23), Hispanic or Latinx (*n* = 8), African American (*n* = 2), Asian (*n* = 2), more than one (*n* = 15), or not fitting into the above races or chose not to answer (*n* = 15).

@@

Using the 1.5-hour videos, our team annotated the periods when the infant was restrained or unrestrained using Datavyu software (https://datavyu.org/). Restrained and unrestrained was defined as a binary category, meaning that all times were coded as one of the two categories. A period was considered missing if the infant went out of view; missing periods were excluded from further analysis. Restraint was defined as when infants could not initiate or control their own movements, which included being held or carried by caregivers, or being restrained in devices that constrain them from changing body position. Unrestrained periods, in contrast, were when infants could freely change their body position or locomote. For example, an infant was considered restrained when strapped into a high chair seat, but not when they were placed in a crib and free to change position. Each video file was coded by two coders: a primary coder annotated the whole 1.5-hour video, and an independent reliability coder annotated the first 30 minutes of each video. Inter-rater reliability was calculated as the proportion of video frames where the two coders coded the same label. The human annotation reached a high agreement with an overall accuracy of `r apa_num(stats_human[1,1]*100)`%, ranging from `r apa_num(stats_human[1,2]*100)`% to `r apa_num(stats_human[1,3]*100)`%, and an average kappa of `r apa_num(stats_human[1,4])`, ranging from `r apa_num(stats_human[1,5])` to `r apa_num(stats_human[1,6])`.

@@

We aggregated raw IMU data into motion features within sliding windows of *n* seconds, where the window length *n* varied from 4 s to 16 s and 30 s. Motion features are a set of statistics to describe a window of data that serve as predictors for the machine learning model. For each window, we computed statistical features—mean, median, minimum, 25th percentile, 75th percentile, maximum, skewness, kurtosis, standard deviation, and sum—for all six signals across the four sensors, resulting in 240 features. We then added 196 additional features based on cross-sensor and cross-orientation metrics such as correlations and pairwise differences, resulting in 436 features per window.

We started the windowing process from the synchronization point and slid the window every *n*/2 seconds. Consequently, each time point spaced every *n*/2 seconds contains 436 motion features, aggregated from the *n*/2 seconds before and after that point. For model training and testing, we included only windows where the infant remained in a single status for over 90% of the time and labeled the window with that status.

The third step was to train and validate supervised machine learning models that classify the motion features into restrained status. We applied the random forest algorithm (Breiman, 2001) in modeling using the *DecisionTree* package (Sadeghi et al., 2022) in Julia. We used a leave-one-out cross-validation approach, where we held out one session at each iteration as testing set and trained the model using the other sessions, to evaluate how well the model generalized to unseen data. After validation, an optimal window length was picked[,]{.underline} and a final model was trained using all the human annotations to process whole-day sensor data of all sessions.

## Reproducibility and Data Sharing

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). This reproducible manuscript, including the model evaluation, figures, and statistical analyses, was made with Quarto version 1.5.57 (Allaire et al., 2024) using R version 4.5.0 (RStudioTeam, 2024) with the R packages tidyverse [@tidyverse], flextable [@flextable], and papaja [@papaja], as well as the apa quarto template (Schneider, n.d.). Human annotation results, which were used for model training and validation, together with processing scripts are publicly shared on OSF (https://osf.io/a698g/overview?view_only=221d1f2bc8d34b5d9838354224d6da8f).

@@

We first compared model performance to the human annotations of the 1.5-hour video recorded period across all 142 sessions using leave-one-out cross-validation approach. We calculated performance metrics including accuracy, kappa, sensitivity, positive predictive value, F1, and the correlation of overall estimate time of restraint between the model and human annotation. We also examined prediction biases model and how they vary across window lengths. Second, we applied the model with the optimal window length to process infants’ full-day sensor data and examined the model’s convergent validity by replicating a previously-reported effect: that restraint time decreases with age [@Carson2022n].

@@

```{r tbl-indexbetweenmodels}
#| tbl-cap: Performance metrics according to window length (4 s, 16 s, and 30 s)
#| ft.align: left
#| apa-note: F, df, and p value indicate one-way ANOVA on each metric comparing across the three window lengths.

@@

```{r fig-prevalencegap}
#| fig-cap: Agreement between model-predicted and human annotated prevalence of restraint
#| fig-width: 8
#| fig-height: 4
#| apa-note: Prevalence is defined as the proportion of windows predicted (or annotated) as restrained among all windows. (A) shows the correlation of prevalence between model prediction and human annotation. The black dashed identity line indicates perfect agreement. Each circular symbol represents one session for a specific window length (denoted by color). (B) shows the difference of prevalence between human annotation and model prediction (i.e., prediction error). A negative prediction error indicates the model estimated less amount of time infants were restrained compared to human annotation. The horizontal line in the box denotes the median; the boarder lines of the box denotes the first (Q1) quartile and the third (Q3) quartile; and the lines extended denotes the min and max data points within the range of 1.5 IQR (i.e., Q3-Q1).

@@

@fig-prevalencegap A shows that, at the session level, all window lengths provided reliable estimate of restrained time. We defined prevalence as the proportion of windows classified as restrained among all windows for each session. The prevalence predicted by the model was highly correlated with the prevalence annotated by human coders (*r* = `r apa_num(table_prev[1,4])` for 4 s, *r* = `r apa_num(table_prev[2,4])` for 16 s, and *r* = `r apa_num(table_prev[3,4])` for 30 s). Thus, model predictions are suitable for capturing individual differences in restrained time between infants.

@@

Despite the overall strong accuracy, models using all three window lengths were biased towards predicting that infants were unrestrained rather than restrained. @fig-cibtwctg shows the difference of a sensitivity and PPV between restrained class and unrestrained class within a single window length. Larger difference scores (away from 0) indicate larger asymmetry in predicting one category rather than the other. @fig-cibtwctg A shows that the sensitivity difference for 4 s window was `r apa_num(table_btw_ctg[1,3])`, 95% bootstrapped CI \[`r apa_num(table_btw_ctg[1,4])`, `r apa_num(table_btw_ctg[1,5])`\], for 16 s window was `r apa_num(table_btw_ctg[3,3])`, 95% CI \[`r apa_num(table_btw_ctg[3,4])`, `r apa_num(table_btw_ctg[3,5])`\], and for 30 s window was `r apa_num(table_btw_ctg[5,3])`, 95% CI \[`r apa_num(table_btw_ctg[5,4])`, `r apa_num(table_btw_ctg[5,5])`\]. The negative difference scores for sensitivity across all window lengths indicate that the model was more sensitive in detecting the unrestrained class compared to restrained class regardless of window length. @fig-cibtwctg B shows that the PPV difference for 4 s window was `r apa_num(table_btw_ctg[2,3])`, 95% CI \[`r apa_num(table_btw_ctg[2,4])`, `r apa_num(table_btw_ctg[2,5])`\], for 16 s window was `r apa_num(table_btw_ctg[4,3])`, 95% CI \[`r apa_num(table_btw_ctg[4,4])`, `r apa_num(table_btw_ctg[4,5])`\], and for 30 s window was `r apa_num(table_btw_ctg[6,3])`, 95% CI \[`r apa_num(table_btw_ctg[6,4])`, `r apa_num(table_btw_ctg[6,5])`\]. The positive differences of PPV across all window lengths indicate that the model’s prediction of restrained periods was more credible compared to its prediction of unrestrained periods. Notably, while none of the window lengths were free from bias, the 30 s window presented smaller offsets of metric values between restrained and unrestrained detection.

The prediction bias was also evident when examining the model’s overall estimate of restrained time and its difference from human annotation. @fig-prevalencegap B shows that the model predictions consistently underestimated overall restrained time compared to human annotation, despite the high correlation between the two. We defined prediction error as the model predicted prevalence (in units of % of recording time) subtracted by the human annotated prevalence of restrained periods (in units of % of recording time). The 4 s window had a prediction error of `r apa_num(table_prev[1,5]*100)`%, 95% CI \[`r apa_num(table_prev[1,6]*100)`%, `r apa_num(table_prev[1,7]*100)`%\], the 16 s window of `r apa_num(table_prev[2,5]*100)`%, 95% CI \[`r apa_num(table_prev[2,6]*100)`%, `r apa_num(table_prev[2,7]*100)`%\], and the 30 s window of `r apa_num(table_prev[3,5]*100)`%, 95% CI \[`r apa_num(table_prev[3,6]*100)`%, `r apa_num(table_prev[3,7]*100)`%\].

Across the metrics of sensitivity, PPV, and prediction error, 30 s turned out to be the best window length out of the three, which brought about the most balanced performance across restrained and unrestrained classes.

@@

# Declarations
## Funding
This work was supported by National Science Foundation Grant BCS #1941449.

## Conflicts of interest
The authors report no conflicts of interest related to this article.

## Ethics approval
All study procedures were reviewed and approved by the Institutional Review Board at the University of California, Riverside (Protocol HS-15-050).

## Consent to participate
Written informed consent was obtained from all caregivers before the study began.
